\chapter{Results and Discussion}

\section{Prediction Model Results}

The Stage 1 prediction model was evaluated as a continuous regression model.
Its purpose is to estimate daily Fire Weather Index (FWI) values from the
locked set of 34 engineered input features. The selected model is a
\texttt{HistGradientBoostingRegressor}, and its output is
\texttt{predicted\_fwi}, which is later used by the dashboard and by the
high-risk decision layer.

The out-of-fold validation results were RMSE \(=4.111\), MAE \(=2.918\), and
\(R^2=0.835\). RMSE and MAE are measured in FWI units, so they describe the
average scale of prediction error relative to the target fire-danger index.
RMSE is more sensitive to larger errors, while MAE gives a more direct average
absolute error. The \(R^2\) value indicates that the model explains a large
portion of the variation in the FWI target during temporal cross-validation.

The 2025 holdout test results were RMSE \(=7.259\), MAE \(=5.098\), and
\(R^2=0.819\). This holdout period is important because it represents a later
seasonal period that was not used as part of the training data. The holdout
performance is weaker than the out-of-fold error values, but the \(R^2\) value
still shows that the model captures much of the FWI pattern on unseen seasonal
data.

Operationally, these results suggest that Stage 1 provides a useful continuous
risk signal, but not a perfect prediction of official FWI. The model can
represent broad fire-weather behaviour, yet error remains possible,
especially under high-variation or unusual weather conditions. This is why the
system does not rely on the regression value alone as the final operational
decision.

The continuous \texttt{predicted\_fwi} value remains valuable for dashboard
decision support. It allows operators to see not only whether a day is treated
as high risk, but also how close the estimated conditions are to the
high-risk boundary. This makes the model output easier to interpret than a
binary label alone.

\section{High-Risk Classification Results}

Stage 2 was evaluated as a high-risk classification layer. It uses a
\texttt{RandomForestClassifier} with four input features:
\texttt{predicted\_fwi}, \texttt{rh}, \texttt{ws}, and
\texttt{fuel\_drying\_rate}. The target is \texttt{target\_ge\_35}, where a
day is considered high risk when \( \mathrm{FWI} \geq 35 \). The stored
probability threshold for the classifier is 0.10.

On the test set, Stage 2 achieved precision \(=0.849\), recall \(=0.900\),
F1-score \(=0.874\), and accuracy \(=0.929\). The confusion matrix contains
45 true positives, 8 false positives, 5 false negatives, and 126 true
negatives. These values show that the classifier identifies most high-risk
days while keeping the number of false alarms at a manageable level for a
prototype decision-support system.

Recall is especially important in this application because a missed high-risk
day can have operational consequences. Precision is also important because too
many false alarms can reduce operator trust and create unnecessary monitoring
load. The F1-score provides a balanced view of precision and recall, while
accuracy is useful but not sufficient by itself because high-risk cases are
less frequent than normal days.

The comparison results show the benefit of the stacked approach
for operational safety. A regression-only decision missed 37 high-risk days on
the 2025 holdout, while the stacked approach reduced missed high-risk days to
6. This does not eliminate all decision risk, but it demonstrates why Stage 2
is useful as a safety-oriented layer above the regression backbone.

The classification result should therefore be interpreted as decision support,
not as a guarantee. Stage 2 improves the handling of high-risk cases near the
decision boundary, but false positives and false negatives remain possible.
For this reason, FireWatch presents the probability, decision reason, and
system context to the operator rather than hiding the decision process behind a
single unexplained label.

\section{Fire and Smoke Detection Results}

The visual detection module uses the active YOLOv8 \texttt{best4.pt} model as
the final fire/smoke detector. The model supports two detection classes: fire
and smoke. Runtime label handling normalises Fire, fire, flame, and flames to
\texttt{fire}, and Smoke or smoke to \texttt{smoke}. Unknown labels are handled
safely rather than being forced into the fire class.

The default detection confidence is configured through
\texttt{MONITORING\_YOLO\_CONF} with a default value of 0.4, and the default
image size is 640. YOLOv8 is suitable for this role because it localises
visible fire or smoke regions using bounding boxes rather than only assigning
an image-level label. This makes the output more useful for live overlays,
snapshot review, and operator interpretation.

Detection alerts preserve operational evidence in the dashboard workflow.
When detections are stored, the alert record can include the fire/smoke label,
confidence value, source camera or device, timestamp, detection list, and
snapshot or evidence reference when available. Tests also confirm that the
detection alert path is separated from the prediction history path, so a
camera alert does not create a wildfire risk prediction run.

The current report treats the detection module as validated through
integration and functional testing rather than through a full benchmark
evaluation. Confirmed tests cover label normalisation, unknown-label handling,
alert persistence, source filtering, alert summaries, latest-alert retrieval,
read/unread state, soft deletion, and separation from the prediction pipeline.
Because no formal benchmark evaluation for \texttt{best4.pt} is available in
the current project evidence, mAP, precision, recall, and dataset-size metrics
are not reported here.

The detector should be interpreted as a prototype visual confirmation module,
not as a certified field detector. False positives from bright light, glare,
reflections, or visually similar objects remain a known limitation. More
advanced robustness methods, such as multi-frame confirmation, glare
filtering, bounding-box area filtering, and extended field validation, are not
implemented in the current prototype.

\section{Dashboard and Operational Behaviour}

The dashboard demonstrates how the FireWatch workflow can be used by an
operator as a decision-support interface. Instead of presenting prediction,
detection, alerts, and system information as separate tools, the dashboard
brings these outputs into a single operational view. This is important because
wildfire monitoring requires both early risk awareness and rapid review of
visual evidence when suspicious fire or smoke appears.

The Overview / Live Monitoring view acts as the main operator-facing area. It
combines the current risk decision, recent weather and system status, and live
monitoring feeds. This view supports quick situational awareness without
requiring the operator to move through several independent tools before
understanding the current state of the system.

The Detection Alerts view provides persistent review of fire and smoke alerts.
Alerts can preserve the detected label, confidence value, source, timestamp,
and snapshot or evidence reference when available. The review workflow also
supports read/unread state and soft deletion, which makes the alert list more
useful for repeated monitoring sessions. This behaviour is important because
visual detections should remain available for later inspection rather than
disappearing after a page refresh or backend restart.

The Risk Map provides spatial context for the Karabük monitoring area and
supports planning-oriented interpretation of historical burned-area and risk
context. It is used as an operator awareness view rather than as a direct
replacement for the prediction model. The System and Model view summarizes
model, feature, and configuration information, while the System Flow view
explains the end-to-end workflow visually. Together, these views make the
prototype easier to interpret as a complete monitoring system.

The main result from the dashboard layer is therefore integration. The
operator can see risk estimates, live monitoring evidence, alert history,
system status, and map context in one interface. This does not make the system
a production emergency platform, but it demonstrates how the individual
prediction and detection modules can be turned into a practical regional
monitoring workflow.

\section{Testing and Validation}

Testing and validation focused on software correctness, integration behaviour,
and persistence rather than official field certification. The confirmed backend
test coverage includes API behaviour, feature and inference checks, scheduler
configuration, scheduled-run persistence, monitoring alerts, and detection
alert state management. A final frontend build result is not reported in this
chapter because no confirmed build record was documented for this reporting
stage.

Scheduler-related tests confirm that the configured scheduled run slots are
registered and that a scheduled risk run can write a persisted
\texttt{run\_history} record. This matters because the scheduler should not be
only a temporary console process; scheduled decisions need to become part of
the same operational history that the dashboard can later display and review.

Database and persistence tests confirm that manual and scheduled prediction
outputs are stored through the SQLite-backed history path. The runtime database
is documented as \texttt{backend/outputs/karabuk\_fwi.db}, which is local
runtime/demo data and is ignored by Git. This policy keeps operational state
out of version control while allowing the local prototype to preserve run
history and alert review state during normal use.

Detection alert testing confirms the main behaviours required by the Detection
Alerts dashboard. The tests cover alert creation, source filtering,
latest-alert retrieval, summary aggregation, detail lookup, read/unread state,
mark-all-read behaviour, marking alerts unread again, soft deletion,
idempotent deletion, snapshot readiness, and preservation of detection labels.
They also confirm that detection alerts remain separated from prediction
history, so camera evidence does not accidentally create a wildfire risk run.

These tests validate that the prototype components are connected in a
controlled and reviewable way. They do not prove field performance under all
weather, camera, lighting, or drone conditions. For this reason, the results
should be interpreted as software and integration validation for an academic
prototype, not as certification for operational emergency deployment.

\section{Limitations}

The Stage 1 regression model still has prediction error. Although the
out-of-fold and holdout results show that the model captures much of the FWI
pattern, difficult years, unusual weather combinations, and high-variation
days can still produce larger errors. The predicted FWI value should therefore
be interpreted as a decision-support estimate rather than a direct substitute
for official fire-danger products.

The Stage 2 classifier improves high-risk decision support, but it does not
remove the possibility of false positives or false negatives. False positives
can increase monitoring load, while false negatives are more serious because
they may reduce attention during genuinely dangerous conditions. This is why
the dashboard presents supporting information such as probability and decision
reason instead of treating the binary high-risk output as an unquestionable
decision.

The visual detection module also has limitations. The current report does not
include formal benchmark metrics such as mAP, precision, recall, or evaluation
dataset size for \texttt{best4.pt}. The detector can identify fire and smoke
labels in the prototype workflow, but false positives may occur from bright
light, glare, reflections, or visually similar objects. Multi-frame
confirmation, glare filtering, bounding-box area filtering, and extended field
validation are not implemented in the current prototype.

The persistence layer is intentionally lightweight. SQLite is suitable for a
single-operator local prototype and for demonstration purposes, but it is not
intended as the final database for a multi-user production deployment. A
production system would likely require a server-based database, stronger
access control, backup strategy, and deployment monitoring.

The drone-ready layer should also be understood carefully. FireWatch supports
mock, Tello-oriented configuration, demo patrol behaviour, stream integration,
and safety gates, but it is not a fully autonomous drone deployment. Full
waypoint patrol, return-to-home logic, battery-aware mission planning,
regulatory flight approval, and production safety validation remain outside
the current project scope. Finally, the repository stores the processed FWI
target used for modelling, while the raw external CEMS/Copernicus download
workflow is treated as part of data preparation rather than as a runtime
dependency.

\section{Discussion}

The results answer the research question by showing that a drone-ready
wildfire monitoring prototype can combine daily risk prediction, visual
fire/smoke detection, persistent alert management, and dashboard-based
decision support for the Karabük region. Each part contributes a different
kind of evidence. Prediction provides early risk awareness, detection provides
visual confirmation, persistence preserves operational history, and the
dashboard turns these outputs into a reviewable workflow.

The two-stage prediction design is useful because it separates continuous risk
estimation from high-risk decision support. Stage 1 provides a numerical FWI
estimate that helps operators understand the severity of conditions. Stage 2
then adds a safety-oriented layer focused on reducing missed high-risk cases
near the decision boundary. This combination is more informative than a single
binary output because it preserves both risk intensity and operational
classification.

The detection module adds an important complementary role. Weather-based risk
prediction cannot confirm whether a fire or smoke is already present, while a
camera detector alone does not explain whether the broader environmental
conditions are dangerous. By keeping prediction and detection separate but
visible in the same dashboard, FireWatch supports both preparedness and visual
evidence review.

Persistence is also central to the system value. Prediction runs and detection
alerts are stored so that decisions and alerts can be reviewed after the
moment of creation. Read/unread state, soft deletion, latest-alert retrieval,
and source filtering make the Detection Alerts workflow closer to an operator
review process rather than a temporary stream of messages.

Overall, FireWatch is strongest as a regional decision-support prototype. It
does not claim official emergency integration, field certification, or
production autonomous drone operation. However, the results demonstrate a
feasible architecture for combining machine learning prediction, computer
vision detection, local persistence, scheduling, and dashboard monitoring in a
Karabük-focused wildfire monitoring system.
